---
title: "BMI Distribution in Scotland from 2008-2012"
author: "2772011M, 2840466A, 2782951G, 2662654J"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
---

```{r}
#| echo: false
#| warning: false

library(tidymodels)
library(readr)
DAProject8 <- read_csv("DAProject8.csv")
```

# Introduction {#sec-intro}

Obesity health concerns are a major issue in Scotland impacting the healthy well-being of individuals across the country. The Body Mass Index (BMI) is a measure to assess weight-related health issues and is used widely across the world to provide insight into the population's health. The Scottish Health Survey (2008-2012) demonstrates how social-economic and lifestyle factors may affect the BMI distribution of the Scottish Population. The survey (displayed in the `DAProject8` dataset) uses factors such as `Age`, `Sex`, `Employment` status, `Veg` intake, `Fruit` intake and more to understand the relationship between social-economic, lifestyle factors and health outcomes of the Scottish population. Here, we shall analyse these relationships and conclude whether there has been a difference in BMI distribution in Scotland over the given years.

@sec-exploratory consists of the Exploratory Analysis of the survey's data and explores the potential trends between social-economic and lifestyle factors and the BMI, and whether there are any differences in the BMI over the given years.

@sec-FA contains the results from fitting a multiple regression model to the data, as well as the assessment of the model assumptions.

The concluding remarks will be given in @sec-conclusion.

# Exploratory Data Analysis {#sec-exploratory}

In this section , we will explore the distribution of BMI from our `DAProject8` dataset.

Here, we summarise BMI patterns and visualise distributions to investigate changes over time and differences by another factor in diet (specifically `Fruit` and `Veg` intake). This will allow us to inform our Formal Analysis in @sec-FA, by checking **normality** and **variance assumptions**.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: setup

library(tidyverse)
library(gt)

DAProject8 <- DAProject8 %>%
  mutate(
    AgeGroup = as.factor(AgeGroup),
    Sex = as.factor(Sex),
    Employment = as.factor(Employment),
    Veg = as.factor(Veg),
    Fruit = as.factor(Fruit),
    Year = as.factor(Year)
  )
```

## Summary Statistics

@tbl-bmi-summary depicts the **interquartile range**, **overall range** and **central tendency** of the BMI in the time frame.

```{r}
#| echo: false
#| eval: true
#| label: tbl-bmi-summary
#| tbl-cap: "Overall Summary Statistics of BMI"

bmi_summary <- DAProject8 %>%
  summarise(
    "Min" = min(BMI, na.rm = TRUE),
    "Q1" = quantile(BMI, 0.25, na.rm = TRUE),
    "Median" = median(BMI, na.rm = TRUE),
    "Mean" = mean(BMI, na.rm = TRUE),
    "Q3" = quantile(BMI, 0.75, na.rm = TRUE),
    "Max" = max(BMI, na.rm = TRUE)
  ) %>%
  gt() %>%
  fmt_number(decimals = 2) %>%
  cols_label(
    Min = "Minimum",
    Q1 = "1st Quartile",
    Median = "Median",
    Mean = "Mean",
    Q3 = "3rd Quartile",
    Max = "Maximum"
  )
bmi_summary
```

From @tbl-bmi-summary, we can see that the interquartile range is significantly narrower than the overall range from the minimum to maximum values observed, suggesting that there may be potential outliers present in the data.

@tbl-bmi-year tracks changes in the **mean**, **median**, **standard deviation**, and s**ample size** over our time frame.

```{r}
#| echo: false
#| eval: true
#| label: tbl-bmi-year
#| tbl-cap: "Mean, Median, and Standard Deviation of BMI by Year"

bmi_by_year <- DAProject8 %>%
  group_by(Year) %>%
  summarise(
    "Mean_BMI" = mean(BMI, na.rm = TRUE),
    "Median_BMI" = median(BMI, na.rm = TRUE),
    "SD_BMI" = sd(BMI, na.rm = TRUE),
    "N" = n()
  ) %>%
  gt() %>%
  fmt_number(decimals = 2) %>%
  cols_label(
    Mean_BMI = "Mean",
    Median_BMI = "Median",
    SD_BMI = "Std. Dev",
    N = "Sample Size"
  )
bmi_by_year
```

As we can see, @tbl-bmi-year shows that both the mean and median BMI remains between 27 and 28 across 2008-2012. Similarly, standard deviations remain approximately constant, indicating no clear trend over time but consistent variability.

## BMI Distribution and Normality

@fig-bmi-hist displays the **overall BMI distribution between 2008 and 2012**.

```{r}
#| echo: false
#| eval: true
#| label: fig-bmi-hist
#| fig-cap: "Histogram of BMI Distribution (2008-2012)"

bmi_hist <- ggplot(DAProject8, aes(x = BMI)) +
  geom_histogram(binwidth = 2, fill = "dodgerblue4", color = "white") +
  labs(title = "Distribution of BMI in Scotland (2008-2012)",
       x = "BMI", y = "Count")
bmi_hist
```

As we can see in @fig-bmi-hist, the distribution appears slightly right-skewed, suggesting potential non-normality. A peak is observed near 25--28 BMI score.

@fig-bmi-qq assesses **normality** with a Q-Q plot.

```{r}
#| echo: false
#| eval: true
#| label: fig-bmi-qq
#| fig-cap: "Q-Q Plot of BMI for Normality Assessment"

qqnorm(DAProject8$BMI, main = "Q-Q Plot of BMI")
qqline(DAProject8$BMI, col = "firebrick1")
```

In @fig-bmi-qq, deviation at the tails - particularly at the upper tail - confirms skewness, suggesting a transformation (such as a log-transform) might be needed for our analysis in @sec-FA.

## BMI Over Time

@fig-bmi-year shows our BMI distribution by `Year`.

```{r}
#| echo: false
#| eval: true
#| label: fig-bmi-year
#| fig-cap: "Boxplot of BMI by Year"

bmi_year_box <- ggplot(DAProject8, aes(x = Year, y = BMI, fill = Year)) +
  geom_boxplot() +
  scale_fill_manual(values = c("2008" = "lightblue", 
                               "2009" = "lightgreen", 
                               "2010" = "lightpink", 
                               "2011" = "lightcoral", 
                               "2012" = "lightgoldenrodyellow")) +
  labs(title = "BMI Distribution by Year",
       x = "Year", y = "BMI", fill = "Year")
bmi_year_box
```

As displayed in @fig-bmi-year, our distributions appear similar throughout, with some outliers above 50 appearing every year, indicating again, that there are no major change over 2008-2012.

## Differences Observed by Diet

@fig-bmi-diet facets BMI by `Veg` and `Fruit` intake across years, which allows us to assess diet effects over time.

```{r}
#| echo: false
#| eval: true
#| label: fig-bmi-diet
#| fig-cap: "Boxplot of BMI by Year and Diet"

DAProject8 <- DAProject8 %>%
  mutate(Veg_Fruit = interaction(Veg, Fruit, sep = "/"))

bmi_diet_box <- ggplot(DAProject8, aes(x = Year, y = BMI, fill = Veg_Fruit)) +
  geom_boxplot() +
  facet_grid(Fruit ~ Veg) + 
  scale_fill_manual(values = c("Yes/Yes" = "green3", 
                               "Yes/No" = "lightgoldenrod1", 
                               "No/Yes" = "lightskyblue", 
                               "No/No" = "firebrick3")) +
  labs(title = "BMI by Year, Vegetable, and Fruit Intake",
       x = "Year", y = "BMI", fill = "Veg/Fruit")
bmi_diet_box
```

As can be observed in @fig-bmi-diet, the BMI distributions remain similar when `Veg` intake is or is not present, no matter what the `Fruit` intake is. We can see that there is potentially a slightly higher BMI associated with both `Veg` and `Fruit` being present in the diet, as the maximum values are generally slightly higher compared to our other combinations of diet.

# Formal Analysis {#sec-FA}

We start by fitting the **full multiple regression model** containing **all** explanatory variables. The full model can be written as:

$$ \begin{align}
y_{i} &= \alpha + \beta_{AgeGroup} \cdot \mathbb{I}_{AgeGroup}(x) + \beta_{Sex} \cdot \mathbb{I}_{Sex}(x) + \beta_{Employment} \cdot \mathbb{I}_{Employment}(x) + \\ &  \quad \beta_{Veg}\cdot \mathbb{I}_{Veg}(x)+ \beta_{Fruit} \cdot \mathbb{I}_{Fruit}(x) + \beta_{Year} \cdot \mathbb{I}_{Year}(x) + \epsilon_i, \\
& ~~~~ \epsilon_i \sim N(0, \sigma^2), ~~~~ i=1,\ldots,n \\
\end{align}$$

where:

-   $\alpha$ is the intercept of the regression line for the baseline of `Fruit` and `Veg` being 'no'.

-   $\beta_{AgeGroup}$ is the intercept of the regression line for the specified `Age` group.

-   $\mathbb{I}_{AgeGroup}(x)$ is an indicator function indicating the `Age` group.

    -   $\mathbb{I}_{AgeGroup}(x)$ is an indicator function indicating the `Age` group.

-   $\beta_{Sex}$ is the intercept of the regression line for the specified `Sex`.

    -   $\mathbb{I}_{Sex}(x) = \begin{cases} 1&\mbox{for male} \\ 0&otherwise \end{cases}$

-   $\beta_{Employment}$ is the intercept of the regression line for the specified `Employment` status.

    -   $\mathbb{I}_{Employment}(x)$ is an indicator function indicating the `Employment` status.

-   $\beta_{Veg}$ is the intercept of the regression line for the specified `Veg` intake.

    -   $\mathbb{I}_{Veg}(x) = \begin{cases} 1&\mbox{for yes} \\ 0&otherwise \end{cases}$

-   $\beta_{Fruit}$ is the intercept of the regression line for the specified `Fruit` intake.

    -   $\mathbb{I}_{Fruit}(x) = \begin{cases} 1&\mbox{for yes} \\ 0&otherwise \end{cases}$

-   $\beta_{Year}$ is the intercept of the regression line for the `Year`.

    -   $\mathbb{I}_{Year}(x)$ is an indicator function indicating the `Year`.

-   $\epsilon_i$ represents normally distributed errors.

## Stepwise Regression

To refine the model, since there are so many variables, we apply **stepwise regression** with **backward selection** in order to reduce the model based on the **Akaike Information Criterion (AIC)**. In doing this, we aim to identify the best-fitting model.

```{r}
#| warning: false
#| echo: false
#| output: false

library(ggplot2)
library(tidyverse)
library(gt)
library(MASS)
library(moderndive)
library(broom)
library(knitr)

#regression model
bmi_model <- lm(BMI ~ AgeGroup + Sex + Employment + Veg + Fruit + Year, data = DAProject8) 
bmi_model

#Stepwise selection using backward elimination
step_model <- stepAIC(bmi_model, direction = "backward", trace = FALSE)
AIC <- step_model$anova
```

The stepwise selection process determines whether all predictors should be included or if a reduced model provides a better fit. In this case, the final model has the lowest AIC of 82517.56, thus is the chosen model.

The final regression model results are summarised in @tbl-regtable.

```{r}
#| echo: false
#| label: tbl-regtable
#| tbl-cap: Estimates of the regression model coeeficients

library(moderndive)

# Display final selected model coefficients
gt(get_regression_table(step_model)) |> fmt_number(decimals = 3)
```

@tbl-regtable provides estimates of the regression model coefficients, showing which factors significantly impact BMI. As displayed, BMI increases with age, with all age groups showing statistically significant positive estimates compared to the reference group (18-24). The largest increase peaks in the 55-64 age group in BMI and then decreases in the 75+ group. With respect to employment status, the overall trend, is when you are in full-time education your BMI tends to be lower than if you are permanently unable to work which illustrates an increase in 1.178 in BMI. Both looking after home/family and looking for/intending to look for paid work are not statistically significant in influencing BMI. Also with respect to lifestyle factors, consuming the recommended vegetable consumption significantly decreases the BMI, which emphasises the importance in vegetables in weight management.

## Model Assumptions

To ensure that regression assumptions hold, we analyse residuals depicted in @fig-residassump below.

```{r, echo=FALSE}
#| echo: false
#| label: fig-residassump
#| fig-cap: "Residual Diagnostic Plots for Stepwise Model"
#| fig-width: 10
#| fig-height: 6

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(step_model)
par(mfrow = c(1, 1))
```

The residual plots help check for **normality**, **homoscedasticity**, and **potential outliers**, all of which seem satisfied.

To confirm the validity of model assumptions, we visualise the **normality of the residuals** in @fig-normalresid below.

```{r}
#| echo: false
#| label: fig-normalresid

# Plot residuals
library(ggplot2)
ggplot(data = data.frame(residuals = residuals(step_model)), aes(x = residuals)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Count")
```

@fig-normalresid verifies that the residuals are approximately normally distributed as it is bell-shaped and centred at zero.

## Interaction Effects

We test for interactions between `Sex` and `Employment`, and `Veg` and `Fruit` intake to explore whether dietary factors interact with employment status or gender.

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false

library(gt)
library(dplyr)

interaction_model <- lm(BMI ~ AgeGroup * Sex + Employment * Veg + Fruit + Year, data = DAProject8)

anova_result <- anova(step_model, interaction_model)
```

```{r}
#| echo: false
#| label: tbl-ANOVA
#| tbl-cap: ANOVA compsrison of our Step Model against the Interaction Model.

anova_df <- as.data.frame(anova_result)

anova_df %>%
  gt() %>%
  tab_header(
    title = "ANOVA Comparison of Models",
    subtitle = "Step Model vs. Interaction Model"
  ) %>%
  cols_label(
    `Res.Df` = "Residual DF",
    RSS = "Residual Sum of Squares",
    Df = "Degrees of Freedom",
    `Sum of Sq` = "Sum of Squares",
    F = "F Statistic",
    `Pr(>F)` = "p-value"
  ) %>%
  fmt_number(
    columns = c(RSS, `Sum of Sq`, F),
    decimals = 2
  ) %>%
  fmt_number(
    columns = `Pr(>F)`,
    decimals = 4
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14
  )
```

As we can observe in @tbl-ANOVA, the ANOVA test assesses whether including interaction terms significantly improves model fit.

## Summary of Findings

-   Certain socio-economic and lifestyle factors significantly impact BMI.
-   Interaction effects between dietary intake and employment status were explored.
-   Stepwise selection refined the model for better interpretability.
-   Diagnostic plots confirm that regression assumptions hold.
-   Residual analysis supports model validity.

The next section will provide concluding remarks based on these findings.

# Conclusions {#sec-conclusion}

As we can see from @tbl-bmi-year and @fig-bmi-year the BMI remained somewhat stable between years 2008-2012, with no significant yearly changes thus time did not have a significant impact on BMI. However, lifestyle and demographic factors do influence BMI as shown by @sec-FA. When it comes to BMI distribution differences by age,gender, socio-economic status or lifestyle factors we can look at @tbl-regtable. Here can we can see that BMI significantly goes on an upward trend as age increases generally. The 55-74 age generally having the largest increase. When it comes Employment we can see that in full time education the BMI decreases significantly. However for employed, self employed and government training seems to have a small but significant impact on increasing BMI. Being perm unable to work has the strongest impact on BMI increase when it comes to employment. Retired also seems to show significant effect in increase of BMI. Looking after home/family and looking/intending looking for work does not show any significant impact on the BMI. When it comes to lifestyle choices it seems that consuming the recommended amount of vegetables of a significant statistical impact on BMI decreasing.
